---
title: "R Workshop 2"
author: "Stephen Johnston"
date: '2019-01-21'
output: html_document
---
```{r}
install.packages(c('tidyverse', 'lubridate', 'devtools', 'googlesheets', 'here'))
devtools::install_github("HakaiInstitute/hakai-api-client-r", subdir='hakaiApi')

library(tidyverse)
library(lubridate)
library(devtools)
library(googlesheets)
library(here)

install.packages("usethis")
library(usethis)
use_git_config(user.name = "stevej-psec", user.email = "stephendanjohnston@gmail.com")

https://github.com/Pelagic-Ecosystems/sockeye-migration.git
```

Source editor is for qriting scripts that we can access later - if written below in the 
console, it's a one time operation shop - liek installing packages

in the Tools tab, under Global Environment - make ssure the "Restore .RData into worksapce 
 at startup" - then our script is whaat saves the data and analysis, rather than the 
 "Environemnt" saved as well - TRYING to be reliant on our Script NOT the Enviro. 

 creating plots with ggplot - if your aes() is placed into the ggplot function then all
 subsequent geomss are based on the associated x and y
 BUT if you put it into the geom_function() then you can place them into the separate geom
 commands
ggplot(data = mpg) + 
  geom_point(aes(x = displ, y = hwy, colour = class)) +
  geom_smooth(aes(x = displ, y = hwy), method = lm)


 Git Hub and Version Control - tracking changes on your data analysis

Commit buttom in the "GIT Environment tab"
getwd()

Packages contain fucnctions that allow us to work with different pre made scripts that are loaded into a specific place


# R MARKDOWN


Facilitates the weaving of plain engligh text, with our coding language so that we can 
follow the prograasion of what we are doing with the code, and makes it intelligable 

This is the sockeye - migration 
https://github.com/Pelagic-Ecosystems/sockeye-migration.git


## Creating Objects in R
```{r setup, warning = FALSE}

```


```{r Simple Arithmetic}
3+5

12/7

5*5

# Assign value to object 
weight_kg <- 55 

weight_lbs <- 2.2*weight_kg # = 121 

# doing slightly more complex calculation -ie. Square Roots 

sqrt(weight_kg) # = 7.42

round(3.14159) # = 3
# but you can include an argument to determine the number of sig figs that we want to round # to
round(3.14159, digits = 3) # = 3.142
# or you can easily just add the number of digits you want to round by putting digits 
round(3.14159, 3) # = 3.142

```
### Vectors and Data Types

This seciton will describe some basic datatypes in R 
Vector is a series of values

```{r}
# numeric vector
weight_g <- c(50,60,65,82)
# character vector
animals <- c("mouse","rat","dog")
# a column is a series of vectors, and a dataframe is a series of columns 

# Atomic Vectors are the base vector - it can only be one type, either - numeric, character, logical (T/F), factor (categorical data), dates

```
Atomic Vectors are the base vector - it can only be one type: 

* numeric
* character
* logical (T/F)
* factor (categorical data)
* dates

A Vector is a data structure in R

Other Data Structures:

* lists
* data frames
* matrices 
* of course vectors

Often you want to convert lists and matricies to data frames or vectors

## Data Frames 

Next we are going to look at the structure of a Data Frame  
```{r}
library(tidyverse)

utils::download.file(url="https://ndownloader.figshare.com/files/2292169", destfile = "read_data/portal_data_joined.csv")

library(here) # This package make working directory an file paths easy

surveys <- read_csv(here("read_data","portal_data_joined.csv"))

surveys 

str(surveys)
dim(surveys)
nrow(surveys)
ncol(surveys)

summary(surveys)

```
Data frame is representation of the vecotrs, where the dataframe has vectors of same length, where matricies can have variable lengths

* A note that certain varaibles are classified as chararcters, rather than factors, ie. Species, would normally be a factor, but listed as a character. 

### Indexing and Subsetting Data Frames

First lets use Square Bracket subsetting

Square brackets are great for definng coordinates to extract data from, but the question is, what happens when your dat frame changes, and th column and row numbers maybe become varied

```{r}
# First define the row coordinate, then he column - always write row, then column
surveys[1,1]

surveys[1,6]

# This defines only which element we want, and will return a data frame
# specifically the first column
surveys[1]
# gives the first row
surveys[1,]
# Gives specific rows adn which column
surveys[1:3,7]

# Specify which columns or rows you don't want with a negative sign (-)
surveys[,-7]
# gets rid os specific columns by number
surveys[-c(5,7,10,12)]

```

## Data Manipulation
Using the Tidyverse to complete data manipulation - better Dialect than BaseR - more intuitive language

* select(): subsettting columns
* filter(): subsetting rows based on conditions
* mutate(): create news column, based oninfomation from other columns
* group_by(): often paired with summarize() - creates groups based on categorical data in a column
* summarize(): creates summary stats on group data 
* arrange(): sort results into a specific order
* count(): gives the number of discrete values 

These functions are much like verbs, indicating what you want to do to data, and they form the sentences that determine what ahppens to the data

A TIBBLE is a tidy data frame... 

```{r}
# Choose specific columns from the DF

select(surveys, plot_id, species_id, weight)
# negative subsetting in the data frame
select(surveys, -record_id)

# Choose Specific Rows from the data frame 

filter(surveys, year == 1995,
       species_id == "NL") 
```

## PIPES

Allow you to chain link together dplyr statements 

Pipe Symbol = %>% or Command + Shift + M


```{r}
# Write multiple arguments in a sentence in dplyr 
surveys %>% 
  filter(weight < 5) %>%
  select(species_id, sex, weight)

surveys_sml <- surveys %>%
  filter(weight < 5) %>%
  select(species_id, sex, weight)

```
Challenge #1 
Using pipes subset the surveys df to include animals collected before 1995 and retain only columns year, sex and weight 
```{r Challenge 1}
challenge1 <- surveys %>%
  filter(year < 1995) %>%
  select(year, sex, weight)
```

```{r Mutate}
surveys %>%
  mutate(weight_kg = weight/1000, weight_kg2 = weight_kg*2)

surveys %>% mutate(mean_weight = mean(weight)) %>%
  View()
# This shows us that the use NAs in calculation just make NAs for everrything

surveys <- surveys %>%
  drop_na(weight) %>%
  mutate(mean_weight = mean(weight))
# this only gives the mean weight for the entire survey df
# better group_by() something specific 
```
Challenge #2
Create new DF from surveys data:
* Contains only species ID column 
* has a new column called hindfoot_half = value that are half the hind foot values
* No NAs and all values are less than 30
```{r}
challenge2 <- surveys %>%
  drop_na(hindfoot_length) %>%
  mutate(hindfoot_half = hindfoot_length/2) %>%
  filter(hindfoot_half < 30) %>%
  select(species_id, hindfoot_half)
  
```

Grouping by Sex
```{r}
surveys %>%
  group_by(sex) %>%
  summarize(mean_weight = mean(weight, na.rm = TRUE)) %>%
              View()

surveys %>% 
  group_by(species_id, sex) %>%
  summarize(mean_weight = mean(weight, na.rm = TRUE)) %>%
  View()

surveys %>% 
  group_by(species_id, sex) %>%
  summarize(mean_weight = mean(weight, na.rm = TRUE),
            min_weight = min(weight, na.rm = TRUE),
            max_weight = max(weight, na.rm=TRUE)) %>%
  View()

# add arrange - does it in ascending order by default 
surveys %>% 
  group_by(species_id, sex) %>%
  summarize(mean_weight = mean(weight, na.rm = TRUE),
            min_weight = min(weight, na.rm = TRUE),
            max_weight = max(weight, na.rm=TRUE)) %>%
  arrange(min_weight) %>%
  View()
# or add 
# arrange(desc(min_weight)) # etc... 

```

```{r}
surveys %>% count(sex) 

# the above code is synonomous with saying 
surveys %>% group_by(sex) %>%
  summarize(count = n())

# sorting it based on the value of the count
surveys %>% count(sex, sort = TRUE)

```
Challenge #3 
1. How many animals were caught in each plot type survey 

2. Usse group_by and summarize to find mean, min and max, of hingfoot_length for each species (using species id col). Also add, number of observations (hint: see ?n)

3. What waas the heaviest animal meeasured in each year? Return the cols: year, genus, species_id, and weight
```{r}
# 1
challenge3.1 <- surveys %>% 
  count(plot_id) 

# 2
challenge3.2 <- surveys %>% group_by(species_id) %>% 
  summarize(count = n(),
            mean = mean(hindfoot_length, na.rm = TRUE),
            min = min(hindfoot_length, na.rm = TRUE),
            max = max(hindfoot_length, na.rm = TRUE)) 

# 3
challenge3.3 <- surveys %>% group_by(year) %>% 
  arrange(weight) %>%
  filter(row_number() %in% c(n())) %>%
  select(year, genus, species_id, weight) %>%
  arrange(year)

# OR
max_weights <- surveys %>% drop_na(weight) %>%
  group_by(year) %>%
  filter(weight == max(weight)) %>%
  select(year, species_id, genus, weight) %>%
  arrange(year) %>% unique()


```

## Export Our DATA

```{r}
write.csv(max_weights, here("write_data", "max_weights.csv"))
```

# DAY TWO

## Tidy Data in Spreadsheets 

The fucntions we use for tidying data are 

* tidyr::spread()
* tidyr::gather()

### spread()

spread() takes three principle arguments:
1. data frame
2. the _key_ column variable will become the new column names
3. the _value_ column variable, which will fill the new column variables

Where key is more likely a descriptor, factor, character, something like "species"
And value is the value associated with that individual - that will be spread from vertical to horizontal 

Using the surveys data set from yesterday - loading it directly from GitHub 

```{r}

surveys <- read_csv(here("read_data","surveys.csv"))

surveys_gw <- surveys %>% 
  drop_na(weight) %>% 
  group_by(species_id,genus) %>%
  dplyr::summarize(mean_wt = mean(weight))

str(surveys_gw)

# NOW we want to spread the dataset out into a wide format

wide_surveys_gw <- surveys_gw %>% spread(key = species_id, value = mean_wt) 

```
Often we want to have our species as a goruping variable in one column - this allows for various graphical outputs to be easilt done

## gather()

Gather() takes 4 arguments instead of three:
1. data
2. key
3. value
4. names of columns we use to gill the keuy variable (or drop) 
```{r}
# Now going from WIDE back to WIDE
long_surveys <- wide_surveys_gw %>%
  gather(key = species_id, value = mean_wt)
```
## DFO GSI Example Problem

Some work done directly on Excel, where some things are easier to complete 
Creating Changelog, Data Dictionary, READme file, 

### Create Changelog 
* Update your changelog with any major changes or steps used to manipulate the data 

```{r}
gsi <- read.csv(here("read_data","tidy_gsi.csv"))

tidy_gsi <- gsi %>% select(hakai_id, stock_1 = Stock.1, region_1 = Region.1, prob_1 = Prob.1)

```

```{r}

```












